|

### 2021-12-16 07-Chp-Triangulation.Rmd
Other edits

However, qualitative triangulation faces issues at scale. There is a need to include all evidence to avoid potential confirmation bias,[@dubroff2018] yet as the number of evidence sources increases, it is not possible to narratively compare and contrast across multiple results in any meaningful way. This fact is illustrated by previous exemplars of triangulation only considering a small number of individual results.[@ference2014; @lawlor2016] To overcome this limitation, some attempts have considered the output of a meta-analysis of similarly-designed studies to be a single source of evidence, and have assessed bias in relation to this summary result. While this approach keeps the number of individual sources of evidence manageable, it loses useful information on the specific biases inherent to each result contributing to the summary estimate. For example, while it may be true that all non-randomised studies share some minimal level of bias, competing extents and directions of bias in individual results may be masked by the use of summary estimates. In addition, previous attempts at qualitative triangulation have not systematically assessed the indirectness of the results, that is the differences between the question of interest and the question addressed by a study.

@@

Additionally, both the qualitative and quantitative triangulation exercises used the detailed risk-of-bias assessments performed and presented as part of the systematic review (Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading)). Similarly, risk-of-bias assessments were performed for each new source of evidence presented in this thesis. The risk of bias tools used to assess each result are described in detail in Section \@ref(risk-of-bias), but in summary, RCTs were assessed using the RoB 2 tool;[@sterne2019] NRSI using the ROBINS-I tool;[@sterne2016] NRSE using an early adapted version of the ROBINS-E tool (see Section \@ref(rob-tools-nrse));[@morganr2020; @french2019] and MR studies using the Mamluk et al. tool.[@mamluk2020]

@@

Analysis of lipid levels across both the systematic review (Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading)) and the IPD analysis (Chapter \@ref(ipd-heading)) found extremely weak evidence for an association of any lipid fraction with all-cause dementia (TC - `r obsLipids$Dementia_TC$estimate`; LDL - `r obsLipids$Dementia_LDL$estimate`; HDL - `r obsLipids$Dementia_HDL$estimate`; TG - `r obsLipids$Dementia_TG$estimate`). Similarly, meta-analysis of studies identified by the review which examined a binary hypercholesterolemia provided weak evidence for an association of this risk factor with Alzheimer's disease (`r obsHyperchol$Dementia$estimate`). Finally, a meta-analysis of RCTs of statin use (`r rct_statin_acd`) and of MR analyses of genetic instruments that mimic statin use (`r mrStatin$Dementia$estimate`) found weak evidence of an association, providing indirect evidence on the effect of lipid lowering on this outcome.

In contrast, the meta-analysis of NRSI of statin use found a slight protective effect for the all-cause dementia outcome (`r obsStatins$Dementia$estimate`), though many of the studies in this analysis were found to be at risk of immortal time bias, which would be expected to favour the statin use. Finally, analysis of the association of statin use with all-cause dementia in the CPRD (Chapter \@ref(cprd-analysis-heading)) found evidence of a harmful association (`r anydem_sta_text`), though this is likely to be driven by the substantial confounding by indication identified in the vascular dementia subgroup.

@@

In contrast, the meta-analysis of NRSI of statin use found a slight protective effect on this outcome (`r obsStatins$AD$estimate`), though similar to the all-cause dementia outcome, many of the studies in this analysis were found to be at risk of immortal time bias. However, the CPRD analysis, which attempted to account for this potential bias through the use of time-varying treatment indicator, found weak evidence for an effect (`r probad_sta_text`). The validity of this finding was limited by the potential for differential misclassification on the basis of the exposure in the CPRD analysis.

Finally, only one of the three cohorts included in the IPD analysis (Whitehall II) contained data on Alzheimer's disease. However, a previously published analysis (Tynkkynen _et al._[@tynkkynen2018]) examining the association of lipid levels with Alzheimer's disease in the this cohort was identified by the systenatic review. The results of this previous analysis are incorporated into the results of the meta-analysis reported above, and so the IPD analysis provided no new information on this outcome.

@@

The IPD analysis provided supporting evidence for the absence of an effect of any fraction on vascular dementia, with the sole exception of triglycerides. Raised triglycerides were associated with an increased incidence of vascular dementia across two previously unanalysed cohorts (`r ipd_vasdem`), though there is the potential for uncontrolled confounding due to the limited information available on important covariates. With respect to evidence on treatments for hypertriglyceridemia, examination of fibrates in the CPRD provided weak evidence of an association with vascular dementia (`r vasdem_fib_text`).

@@

1. Define the causal question of interest
1. Identify relevant evidence sources and standardise effect directions
1. Specify idealised version of each study
1. Assess the extent and direction of bias/indirectness in each result
1. Define modifying terms for bias and indirectness in each result
1. Calculate bias-/indirectness-adjusted results and perform meta-analysis

@@

Once the causal question of interest had been defined, relevant individual results were obtained from the data sources described in Section \@ref(tri-data-sources). In a broad sense, this meant extracting results that examined the relationship between the exposure/outcome pair, either directly (e.g. non-randomised studies of LDL-c levels) or indirectly (e.g. non-randomised studies of statins, or Mendelian randomisation studies using genetic instruments for LDL-c levels).

@@

Once the set of relevant results have been identified, an idealised study for each is described
An idealised study can be viewed as a replicate of the study producing the result of interest but with all sources of bias removed. The risk of bias is then assessed against this idealised study, while the causal question addressed by each idealised study is compared with the target causal question to define the indirectness of each result.

@@

The assessment of bias in each result is discussed in Section \@ref(tri-data-sources) above.For the sake of consistency and ease of computing, I mapped the different acceptable risk-of-bias judgements in each tool to a harmonised set: "Low", "Moderate", and "High". The exact mapping can be seen in Table \@ref(tab:robLevelsMapping-table). Of note, no mapping was performed for the critical risk-of-bias levels present in the tools for non-randomised studies. This is in accordance with current best practice in evidence synthesis, which is to exclude all studies at critical risk of bias from further quantitative synthesis.[@sterne2016]

@@

@@

The predicted direction and type of bias were not recorded for domains at "Low" risk of bias. Finally, for some domains, only one type of bias is allowed. For example, in the confounding domains in the ROBINS-I/ROBINS-E tools, only the "Favours experimental"/"Favours comparator" and "Unpredictable" options are available, meaning that bias in this domain will always be considered additive.

@@

Reasonable values for the position and spread of the prior distributions were informed by a previously reported expert bias elicitation exercise.[@turner2009] Using open data from that study, I calculated the mean values for the mean and variance of the adjustment distributions across all biases considered by that exercise (see Appendix \@ref(appendix-adjustment-values) for more details on how these were calculated). For the purposes of this illustrative analysis and to highlight the generalised nature of the code used to perform the analysis (see Section \@ref(tri-software)), I used the calculated values to inform two scenarios of modifying distributions (Table \@ref(tab:priorsAdd-table)) for additive biases, and a single set for proportional biases. In the first scenario, the relationship between the position assigned to each additive judgement is linear. In the second scenario, the step from "Moderate" to "High" is twice that from "Low" to "Moderate". Across both additive and proportional biases, the variance of the "High" judgement distribution was defined to be greater than that of the "Moderate" judgement.

@@

As an illustrative example, consider the first causal question of the interest in this exercise, the effect of LDL-c in mid-life on Alzheimer's disease. Here, studies of lipids levels in this time period would require minimal adjustment, whereas studies examining statin use (indirect exposure/intervention) in late-life (indirect population) would be adjusted and down-weighted due to reduced relevance to the causal question.

@@

where $\mu_{i\delta}^{\mathrm{B}}$ and $\mu_{i\beta}^{\mathrm{B}}$ refer to the total additive and proportional bias, and  $\mu_{i\delta}^{\mathrm{In}}$ and $\mu_{i\beta}^{\mathrm{In}}$  refer to the total additive and proportional indirectness. Note that in this exercise, the total additive indirectness $\mu_{i\beta}^{\mathrm{In}} = 1$, as all indirectness was defined _a priori_ as being proportional.

The standard error of this estimate is then calculated as:

@@

As an illustrative example of the process using a single result, the bias and indirectness inherent to the estimate of the effect of statin use on Alzheimer's disease in the CPRD (as analysed in Chapter \@ref(cprd-analysis-heading)) are considered here. For this result, there were three domains at greater than low risk of bias: Bias due to confounding, Bias due to definition of the outcome, and Bias in the selection of the reported result (Figure \@ref(fig:biasDirectionSingle)). In Domain 1 (bias due to confounding), the study did not adjust for _Apo_$\mathcal{E}4$ status, which is predicted to mask any true association between statins and Alzheimer's disease (see Section \@ref(cprd-limitations) for a fuller discussion). As such, the bias is predicted to _favour the comparator_ and so the absolute direction of bias is to the right. In Domain 7 (bias in measurement of the outcome), there was a high risk of differential misclassification on the basis of the exposure. Here, a history of statin use may make a diagnosis of vascular dementia more likely than Alzheimer's disease, and so for this outcome, this bias is predicted to _favour the experimental_ (in this case, an intervention). As such, the absolute direction of bias is to the left. In Domain 7 (bias in selection of the reported result), for consistency with the other NRSI, in the absence of a protocol the bias is assumed to be _away from the null_. Given the position of the effect estimate below the null, the absolute direction of bias in this domain is to the left.

@@

Indirectness was intentionally not displayed in the plot above, as these figures are created using `robvis` which is focused on creating visualisations of risk-of-bias data. In addition, I found it more useful to articulate the indirectness of a result in tabular format. With respect to this example result, I judged it to be at low risk in the population and outcome domains (Table \@ref(tab:singleIndirect-table)). However, the result refers to the effect of statins on Alzheimer's disease risk, rather than LDL-c levels directly, putting it at high risk of indirectness in the intervention/exposure domain. Overall, assessing the predicted direction of indirectness was more challenging than assessing the predicted direction of bias. In this case, treatment with statins may result in temporary lowering of lipid levels which could thus underestimate the true effect of LDL-c lowering (indirectness towards the null). In contrast, statins may have an effect on Alzheimer's disease through a mechanism other than their LDL-c lowering effect and so the effect may be overestimated (indirectness away from null). As such, I judged the direction of indirectness in this domain to be unpredictable.

@@

&nbsp;<!----------------------------------------------------------------------->  
(ref:singleIndirect-caption) __Assessment of indirectness for a single example study__

@@

The unadjusted estimate was `r single_unadjusted `. After accounting for the biases and indirectness described, the adjusted result was `r single_adjusted `, representing a minimal shift in the position of the effect estimate but a substantial decrease in precision.

@@

From the data sources identified in Section \@ref(tri-data-sources), I identified 18 results relevant to my first causal question of interest. Most (N =17) were identified via the systematic review,`r ldl_ad_citations` while Chapter \@ref(cprd-analysis-heading) provided an additional estimate from the CPRD analysis.  The extent, type and direction of predicted bias for each result can be seen in the bias-direction plot presented in Figure \@ref(fig:ldlAdBiasDirection), stratified by study design.

<!-- QUESTION Hypercholesterolemia should not be included here, as it is focused on total cholesterol rather than specifically LDL-c, right? -->

@@

The results of the unadjusted and the bias-/indirectness-adjusted (under Scenario 1) are shown in Figure \@ref(fig:fpLdlAd). Following adjustment for bias and indirectness, the observed effect attenuates to the null (unadjusted `r unadj_effect`; adjusted `r adj_effect_scenario1`). Additionally, adjustment for bias and indirectness substantially reduced the heterogeneity between studies (unadjusted $\tau^2$ = `r unadj_tau2`, $I^2$ = `r unadj_I2`; adjusted  $\tau^2$ = `r adj_tau2`, $I^2$ = `r adj_I2`). However, as can been seen from the right panel in Figure \@ref(fig:fpLdlAd), this reduction in heterogeneity between studies is largely achieved via a reduction in the precision of each individual result.

@@

Comparison of the unadjusted and bias-/indirectness-adjusted results for the effect of triglycerides on vascular dementia did not demonstrate a substantial difference (unadjusted `r unadj_effect_vad`; adjusted `r adj_effect_scenario1_vad`; Figure \@ref(fig:fpTgVad)), though again, heterogeneity between the adjusted results was greatly reduced (unadjusted $\tau^2$ = `r unadj_vad_tau2`, $I^2$ = `r unadj_vad_I2`; adjusted  $\tau^2$ = `r adj_vad_tau2`, $I^2$ = `r adj_vad_I2`). Similarly there was minimal difference between the results under the two scenarios of bias (Figure \@ref(fig:fpTgVaDComparison) in Appendix \@ref(appendix-tri-analysis)).

@@

Ideally, these prior distributions would be based on empirical data on the effect of different domains of bias/indirectness on a result. Most meta-epidemiological studies examining the effect of different methodological issues studied only randomised controlled trials,[@amer2021; @page2016] and the estimates of bias they produce may not hold for non-randomised study designs. However, the prior distributions defined here are broadly comparable to previously reported estimates of the impact of bias in non-randomised trials. For example, a previous simulation study estimated that, in the absence of true effect of LDL-c, _Apo_$\mathcal{E}4$ would induce a RR of 1.09 with risk of dementia per 1-SD increase in LDL-c.[@iwagami2021] This is comparable to the mean assigned to the prior log-normal distribution mapped to "Moderate" additive bias ($N(0.09,0.05)$, Table \@ref(tab:priorsAdd-table)).

@@

Finally, it may in fact be more reasonable for domains such as bias due to confounding to adjust on a per-confounder basis rather than mapping to a "Moderate" or "High" extent of bias. In this approach, prior distributions are assigned to each of the confounders pre-specified as part of the ROBINS-I/ROBINS-E tools.[@sterne2016] Then, rather than assigning a domain level distribution based on an arbitrary judgement of how many important confounders are missing, the result is adjusted for each confounder not accounted for in the original analysis.

@@

A final limitation of this analysis is the synthesis of different effect estimates. As discussed in Section \@ref(sys-rev-analysis-overview), hazard ratios are not directly comparable to odds ratios, as they inherently account for person time-at-risk.[@mckenzie2019] As non-randomised studies (NRSI/NRSI) of dementia outcomes report hazard ratios and Mendelian randomisation studies report odds ratios, the resulting synthesis may be biased by the integration of these two different effect measures. This is primarily a concern for common outcomes, as when the outcome is rare, the odds, risk and hazard ratios approximate each other.

@@

A further advantage of this method comes from the ability to specify the prior distributions for each level of bias/indirectness in advance of performing the assessments. Expert elicitation of the extent of bias/indirectness using a panel of methodologists and topic experts is a powerful technique but the point at which it is deployed in the framework is important. In previous attempts at bias-/indirectness-adjusted meta-analysis, the extent of bias in each study was assessed via an elicitation process,[@thompson2011; @turner2009; @wilks2011] during which experts were aware of the results of each analysis. This approach is potentially subject to differential misclassification of the impact of bias/indirectness on the basis of the results, as there is no way to ensure that results at a similar level of bias for confounding (for example) are being adjusted by the same amount across studies. As an illustrative example, if experts are influenced by the knowledge of the effect estimate in a study, then a result at "Moderate" risk of confounding and demonstrating a stronger protective effect may receive greater adjustment than a study at the same risk of confounding but with a more modest effect. This is likely to be particularly problematic where experts have preconceived ideas about the true effectiveness of the intervention.

@@

Future development of this framework should also account for bias at the analysis level in terms of missing evidence.[@zotero-15123] For example, there is at least one known study relevant to the triglycerides/vascular dementia case study that reported a non-significant result but provided insufficient details to be included in the analysis.[@chiang2007] Ideally, a quantitative triangulation framework would further account for this proportional meta-bias (as bias due to missing evidence is most likely to bias away from the null due to publication bias mechanisms), in addition to result-specific biases.

Appreciation that both the risk-of-bias and indirectness of a result should be assessed is growing. Indeed, some existing tools for the assessment of diagnostic test accuracy[@whiting2011] and prediction models[@moons2019] consider indirectness (termed applicability in these tools) alongside the assessment of bias. Future iterations of the core risk-of-bias tools (RoB 2, ROBINS-I, ROBINS-E) could take this into account, though there is a competing argument that indirectness is already handled via tools such as the GRADE framework.[@guyatt2011] In addition, simple steps like harmonising the risk-of-bias judgements across tools and making the direction of bias question mandatory (even if users default to "Unpredictable") would represent an improvement in the tools and encourage users to think about how the biases assessed affect the corresponding result. Similarly, software that implements a risk-of-bias tool should allow for direction/extent of bias question and carefully consider how this data will be exported.

@@

There are two key outputs from this project. The first is the `triangulate` R package. Personal communication with the authors of the original method resulted in the STATA code to implement the bias-/indirectness-adjusted model.[@turner2009] This has since been generalised as part of the `triangulate` R package to enable other users to apply the approach detailed here. The package allows for preprocessing of domain-based risk-of-bias data to correctly identify the direction of bias relative to the effect estimate, use of domain-specific prior distributions, production of bias direction plots (via the tool described in Appendix \@ref(appendix-robvis), calculation of bias/indirectness-adjusted estimates for each result, and synthesis of these adjusted results in a standard random-effects meta-analysis.

>>>>

However, qualitative triangulation faces issues at scale. There is a need to include all evidence to avoid potential confirmation bias,[@dubroff2018] yet as the number of evidence sources increases, it is not possible to narratively compare and contrast across multiple results in any meaningful way. This fact is illustrated by previous exemplars of triangulation only considering a small number of individual results.[@ference2014; @lawlor2016] To overcome this limitation, some attempts have considered the output of a meta-analysis of similarly-designed studies to be a single source of evidence, and have assessed bias in relation to this summary result. While this approach keeps the number of individual sources of evidence manageable, it loses useful information on the specific biases inherent to each result contributing to the summary estimate. For example, while it may be true that all non-randomised studies share some minimal level of bias, competing extents and directions of bias in individual results may be masked by the use of summary estimates. In addition, previous attempts at qualitative triangulation have not systematically assessed the indirectness of the results, namely the differences between the question of interest and the question addressed by a study.

@@

Additionally, both the qualitative and quantitative triangulation exercises used the detailed risk-of-bias assessments performed and presented as part of the systematic review (Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading)). Similarly, risk-of-bias assessments were performed for each new source of evidence presented in this thesis. The risk of bias tools used to assess each result are described in detail in Section \@ref(risk-of-bias), but in summary, randomised controlled trials (RCTs) were assessed using the RoB 2 tool;[@sterne2019] non-randomised studies of interventions (NRSI) using the ROBINS-I tool;[@sterne2016] non-randomised studies of exposures (NRSE) using an early adapted version of the ROBINS-E tool (see Section \@ref(rob-tools-nrse));[@morganr2020; @french2019] and Mendelian randomisation (MR) studies using the Mamluk et al. tool.[@mamluk2020]

@@

Analysis of lipid levels across both the systematic review (Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading)) and the individual participant data (IPD) analysis (Chapter \@ref(ipd-heading)) found extremely weak evidence for an association of any lipid fraction with all-cause dementia (in the systematic review: TC - `r obsLipids$Dementia_TC$estimate`; LDL - `r obsLipids$Dementia_LDL$estimate`; HDL - `r obsLipids$Dementia_HDL$estimate`; TG - `r obsLipids$Dementia_TG$estimate` ). Similarly, meta-analysis of studies identified by the review which examined a binary hypercholesterolemia provided weak evidence for an association of this risk factor with Alzheimer's disease (`r obsHyperchol$Dementia$estimate`). Finally, a meta-analysis of RCTs of statin use (`r rct_statin_acd`) and of MR analyses of genetic instruments that mimic statin use (`r mrStatin$Dementia$estimate`) found weak evidence of an association, providing indirect evidence on the effect of lipid lowering on this outcome.

In contrast, the meta-analysis of statin use found a slight protective effect for the all-cause dementia outcome (`r obsStatins$Dementia$estimate`), although many of the studies in this analysis were found to be at risk of immortal time bias, which would be expected to favour the statin use. Finally, analysis of the association of statin use with all-cause dementia in the CPRD (Chapter \@ref(cprd-analysis-heading)) found evidence of a harmful association (`r anydem_sta_text`). However, this finding is likely to be driven by the substantial confounding by indication identified in the vascular dementia subgroup.

@@

In contrast, the meta-analysis of statin use found a slight protective effect on this outcome (`r obsStatins$AD$estimate`), though similar to the all-cause dementia outcome, many of the studies in this analysis were found to be at risk of immortal time bias. However, the CPRD analysis, which attempted to account for this potential bias through the use of time-varying treatment indicator, found weak evidence for an effect (`r probad_sta_text`). The validity of this finding was limited by the potential for differential misclassification on the basis of the exposure in the CPRD analysis.

Finally, only one of the three cohorts included in the IPD analysis (Whitehall II) contained data on Alzheimer's disease. However, a previously published analysis (Tynkkynen _et al._[@tynkkynen2018]) examining the association of lipid levels with Alzheimer's disease in this cohort was identified by the systenatic review. The results of this previous analysis are incorporated into the results of the meta-analysis reported above, and so the IPD analysis provided no new information on this outcome.

@@

The IPD analysis provided supporting evidence for the absence of an effect of any fraction on vascular dementia, with the sole exception of triglycerides. Raised triglycerides were associated with an increased incidence of vascular dementia across two previously unanalysed cohorts (`r ipd_vasdem`), although there is the potential for uncontrolled confounding due to the limited information available on important covariates. With respect to evidence on treatments for hypertriglyceridemia, examination of fibrates in the CPRD provided weak evidence of an association with vascular dementia (`r vasdem_fib_text`).

@@

1. Define the causal question of interest
1. Identify relevant evidence sources and standardise effect directions
1. Specify an idealised version of each study
1. Assess the extent and direction of bias/indirectness in each result
1. Define modifying terms for bias and indirectness in each result
1. Calculate bias-/indirectness-adjusted results and perform meta-analysis

@@

Once the causal question of interest had been defined, relevant individual results were obtained from the data sources described in Section \@ref(tri-data-sources). In a broad sense, this meant extracting results that examined the relationship between the exposure/outcome pair, either directly (e.g. non-randomised studies of LDL-c levels) or indirectly (e.g. non-randomised studies of statins, or MR studies using genetic instruments for LDL-c levels).

@@

Once the set of relevant results have been identified, an idealised study for each should be described. An idealised study can be viewed as a replicate of the study producing the result of interest but with all sources of bias removed. The risk of bias is then assessed against this idealised study, while the causal question addressed by each idealised study is compared with the target causal question to define the indirectness of each result.

@@

The assessment of bias in each result is discussed in Section \@ref(tri-data-sources) above. For the sake of consistency and ease of computing, I mapped the different acceptable risk-of-bias judgements in each tool to a harmonised set: "Low", "Moderate", and "High". The exact mapping can be seen in Table \@ref(tab:robLevelsMapping-table). Of note, no mapping was performed for the critical risk-of-bias levels present in the tools for non-randomised studies. This is in accordance with current best practice in evidence synthesis, which is to exclude all studies at critical risk of bias from further quantitative synthesis.[@sterne2016]

@@

<!-- QUESTION text -->

@@

The predicted direction and type of bias was not recorded for domains at "Low" risk of bias. Finally, for some domains, only one type of bias is allowed. For example, in the confounding domains in the ROBINS-I/ROBINS-E tools, only the "Favours experimental"/"Favours comparator" and "Unpredictable" options are available, meaning that bias in this domain will always be considered additive.

@@

Reasonable values for the position and spread of the prior distributions were informed by a previously reported expert bias elicitation exercise.[@turner2009] Using open data from that study, I calculated the average values for the mean and variance of the adjustment distributions across all biases considered by that exercise (see Appendix \@ref(appendix-adjustment-values) for more details on how these were calculated). For the purposes of this illustrative analysis and to highlight the generalised nature of the code used to perform the analysis (see Section \@ref(tri-software)), I used the calculated values to inform two scenarios of modifying distributions (Table \@ref(tab:priorsAdd-table)) for additive biases, and a single set for proportional biases. In the first scenario, the relationship between the position assigned to each additive bias judgement is linear. In the second scenario, the step from "Moderate" to "High" is twice that from "Low" to "Moderate". Across both additive and proportional biases, the variance of the "High" judgement distribution was defined to be greater than that of the "Moderate" judgement.

@@

As an illustrative example, consider the first causal question of the interest in this exercise, the effect of LDL-c in mid-life on Alzheimer's disease. Here, studies of lipid levels in this time period would require minimal adjustment, whereas studies examining statin use (indirect exposure/intervention) in late-life (indirect population) would be adjusted and down-weighted due to reduced relevance to the causal question.

@@

where $\mu_{i\delta}^{\mathrm{B}}$ and $\mu_{i\beta}^{\mathrm{B}}$ refer to the total additive and proportional bias, and  $\mu_{i\delta}^{\mathrm{In}}$ and $\mu_{i\beta}^{\mathrm{In}}$  refer to the total additive and proportional indirectness, respectively. Note that in this exercise, the total additive indirectness $\mu_{i\beta}^{\mathrm{In}} = 1$ because all indirectness was defined _a priori_ as being proportional.

The standard error of this estimate ($\mathrm{SE}(\hat{\theta})$) is then calculated as:

@@

As an illustrative example of the process using a single result, the bias and indirectness inherent to the estimate of the effect of statin use on Alzheimer's disease in the CPRD (as analysed in Chapter \@ref(cprd-analysis-heading)) are considered here. For this result, there were three domains at greater than low risk of bias: bias due to confounding, bias due to definition of the outcome, and bias in the selection of the reported result (Figure \@ref(fig:biasDirectionSingle)). In Domain 1 (bias due to confounding), the study did not adjust for _Apo_$\mathcal{E}$ status, which is predicted to mask any true association between statins and Alzheimer's disease (see Section \@ref(cprd-limitations) for a fuller discussion). As such, the bias is predicted to _favour the comparator_ and so the absolute direction of bias is to the right. In Domain 6 (bias in measurement of the outcome), there was a high risk of differential misclassification on the basis of the exposure. Here, a history of statin use may make a diagnosis of vascular dementia more likely than Alzheimer's disease, and so for this outcome, this bias is predicted to _favour the experimental_ (in this case, an intervention). As such, the absolute direction of bias is to the left. In Domain 7 (bias in selection of the reported result), for consistency with the other NRSI, in the absence of a protocol the bias is assumed to be _away from the null_. Given the position of the effect estimate below the null, the absolute direction of bias in this domain is to the left.

@@

Indirectness was intentionally not displayed in the plot above because these figures are created using `robvis` which is focused on creating visualisations of risk-of-bias data. In addition, I found it more useful to articulate the indirectness of a result in tabular format. With respect to this example result, I judged it to be at low risk in the population and outcome domains (Table \@ref(tab:singleIndirect-table)). However, the result refers to the effect of statins on Alzheimer's disease risk, rather than LDL-c levels directly, putting it at high risk of indirectness in the intervention/exposure domain.

Overall, assessing the predicted direction of indirectness was more challenging than assessing the predicted direction of bias. In this case, treatment with statins may result in temporary lowering of lipid levels which could thus underestimate the true effect of LDL-c lowering (indirectness towards the null). In contrast, statins may have an effect on Alzheimer's disease through a mechanism other than their LDL-c lowering effect and so the effect may be overestimated (indirectness away from null). As such, I judged the direction of indirectness in this domain to be unpredictable.

@@

&nbsp;<!----------------------------------------------------------------------->  
(ref:singleIndirect-caption) __Assessment of indirectness for a single example study__ - The target setting for the causal question of interest (effect of midlife LDL-c on Alzheimer's disease) and the idealised version of a single example study are compared. The risk and predicted direction of indirectness resulting from discrepancies between the target setting and idealised study are shown.

@@

The unadjusted estimate of the effect of LDL-c on Alzheimer's disease was `r single_unadjusted `. After accounting for the biases and indirectness described, the adjusted result was `r single_adjusted `, representing a minimal shift in the position of the effect estimate but a substantial decrease in precision.

@@

<!-- TODO Update this paragraph to reflect extra 2 MR studies -->

From the data sources identified in Section \@ref(tri-data-sources), I identified 18 results relevant to my first causal question of interest. Most (N =17) were identified via the systematic review,`r ldl_ad_citations` while Chapter \@ref(cprd-analysis-heading) provided an additional estimate via the CPRD analysis.  The extent, type and direction of predicted bias for each result can be seen in the bias-direction plot presented in Figure \@ref(fig:ldlAdBiasDirection), stratified by study design.

<!-- TODO Double check this reference list is correct! -->

@@

The results of the unadjusted and the bias-/indirectness-adjusted (under Scenario 1) are shown in Figure \@ref(fig:fpLdlAd). Following adjustment for bias and indirectness, the observed effect attenuates to the null (unadjusted `r unadj_effect`; adjusted `r adj_effect_scenario1`). Additionally, adjustment for bias and indirectness substantially reduced the heterogeneity between studies (unadjusted $\tau^2$ = `r unadj_tau2`, $I^2$ = `r unadj_I2`%; adjusted  $\tau^2$ = `r adj_tau2`, $I^2$ = `r adj_I2`%). However, as can been seen from the right panel in Figure \@ref(fig:fpLdlAd), this reduction in heterogeneity between studies is largely achieved via a reduction in the precision of each individual result.

@@

Comparison of the unadjusted and bias-/indirectness-adjusted results for the effect of triglycerides on vascular dementia did not demonstrate a substantial difference (unadjusted `r unadj_effect_vad`; adjusted `r adj_effect_scenario1_vad`; Figure \@ref(fig:fpTgVad)), though again, heterogeneity between the adjusted results was greatly reduced (unadjusted $\tau^2$ = `r unadj_vad_tau2`, $I^2$ = `r unadj_vad_I2`%; adjusted  $\tau^2$ = `r adj_vad_tau2`, $I^2$ = `r adj_vad_I2`%). Similarly there was minimal difference between the results under the two scenarios of bias (Figure \@ref(fig:fpTgVaDComparison) in Appendix \@ref(appendix-tri-analysis)).

@@

Ideally, these prior distributions would be based on empirical data on the effect of different domains of bias/indirectness on a result. Most meta-epidemiological studies examining the effect of different methodological issues studied only randomised controlled trials,[@amer2021; @page2016] and the estimates of bias they produce may not hold for non-randomised study designs. However, the prior distributions defined here are broadly comparable to previously reported estimates of the impact of bias in non-randomised trials. For example, a previous simulation study estimated that, in the absence of a true effect of LDL-c, _Apo_$\mathcal{E}4$ would induce a risk ratio of 1.09 with risk of dementia per 1-SD increase in LDL-c.[@iwagami2021] This is comparable to the mean assigned to the prior log-normal distribution mapped to "Moderate" additive bias ($N(0.09,0.05)$, Table \@ref(tab:priorsAdd-table)).

@@

Finally, for the "bias due to confounding" domain, it may in fact be more reasonable to adjust on a per-confounder basis rather than mapping to a "Moderate" or "High" extent of bias. In this approach, prior distributions are assigned to each of the confounders pre-specified as part of the ROBINS-I/ROBINS-E tools.[@sterne2016] Then, rather than assigning a domain level distribution based on an arbitrary judgement of how many important confounders are missing, the result is adjusted for each confounder not accounted for in the original analysis.

@@

A final limitation of this analysis is the synthesis of different effect estimates. As discussed in Section \@ref(sys-rev-analysis-overview), hazard ratios are not directly comparable to odds ratios because they inherently account for person time-at-risk.[@mckenzie2019] Non-randomised studies (NRSI/NRSI) of dementia outcomes reported hazard ratios and Mendelian randomisation studies reported odds ratios, and as such the resulting synthesis may be biased by the integration of these two different effect measures. This is primarily a concern for common outcomes because the odds, risk and hazard ratios approximate each other when the outcome is rare.

@@

A further advantage of this method comes from the ability to specify the prior distributions for each level of bias/indirectness in advance of performing the assessments. Expert elicitation of the extent of bias/indirectness using a panel of methodologists and topic experts is a powerful technique but the point at which it is deployed in the framework is important. In previous attempts at bias-/indirectness-adjusted meta-analysis, the extent of bias in each study was assessed via an elicitation process,[@thompson2011; @turner2009; @wilks2011] during which experts were aware of the results of each analysis. This approach is potentially subject to differential misclassification of the impact of bias/indirectness on the basis of the results because there is no way to ensure that results at a similar level of bias for confounding (for example) are being adjusted by the same amount across studies. As an illustrative example, if experts are influenced by the knowledge of the effect estimate in a study, then a result at "Moderate" risk of confounding and demonstrating a stronger protective effect may receive greater adjustment than a study at the same risk of confounding but with a more modest effect. This is likely to be particularly problematic where experts have preconceived ideas about the true effectiveness of the intervention.

@@

Future development of this framework should also account for bias at the analysis level in terms of missing evidence.[@zotero-15123] For example, there is at least one known study relevant to the triglycerides/vascular dementia case study that reported a non-significant result but provided insufficient details to be included in the analysis.[@chiang2007] Ideally, a quantitative triangulation framework would further account for this proportional meta-bias (seeing as bias due to missing evidence is most likely to bias away from the null due to publication bias mechanisms), in addition to result-specific biases.

Appreciation that both the risk-of-bias and indirectness of a result should be examined is growing. Indeed, some existing tools for the assessment of diagnostic test accuracy[@whiting2011] and prediction models[@moons2019] consider indirectness (termed applicability in these tools) alongside the assessment of bias. Future iterations of the core risk-of-bias tools (RoB 2, ROBINS-I, ROBINS-E) could take this into account, though there is a competing argument that indirectness is already handled via tools such as the GRADE framework.[@guyatt2011] In addition, simple steps like harmonising the risk-of-bias judgements across tools and making the direction of bias question mandatory (even if users default to "Unpredictable") would represent an improvement in the tools and encourage users to think about how the biases assessed affect the corresponding result. Similarly, software that implements a risk-of-bias tool should allow for direction/extent of bias question and carefully consider how this data will be exported.

@@

There are two key outputs from this chapter. The first is an R package, `triangulate`. Personal communication with the authors of the bias-/indirectness-adjusted meta-analysis approach resulted in the STATA code to implement the adjusted model.[@turner2009] This has since been generalised as part of the `triangulate` R package to enable other users to apply the approach detailed here. The package allows for preprocessing of domain-based risk-of-bias data to correctly identify the direction of bias relative to the effect estimate, use of domain-specific prior distributions, production of bias direction plots (via the tool described in Appendix \@ref(appendix-robvis)), calculation of bias/indirectness-adjusted estimates for each result and synthesis of these adjusted results in a standard random-effects meta-analysis.
